Evolution Reinforces Cooperation with the Emergence of Self-Recognition Mechanisms: an Empirical Study of the Moran process for the Iterated Prisoner's Dilemma

### 2017-10-23 main.tex
Make minor fixes.

% Title
\title{Evolution Reinforces Cooperation with the Emergence of Self-Recognition
       Mechanisms: an empirical study of the Moran process for the iterated
       Prisoner's dilemma}
\author[1]{Vincent Knight}
\author[2]{Marc Harper}
\author[1]{Nikoleta E. Glynatsi}
\author[3]{Owen Campbell}

@@

The Prisoner's Dilemma (PD) \cite{Flood1958} is a fundamental two player game
used to model a variety of strategic interactions. Each player chooses
simultaneously and independently
between cooperation (C) or defection (D). The payoffs of
the game are defined by the matrix $\begin{pmatrix} R & S \\ T & P
\end{pmatrix}$, where $T > R > P > S$ and $2R > T + S$. The PD is a one
round game, but is commonly studied in a manner where the prior outcomes
matter. This repeated form is called the Iterated Prisoner's
Dilemma (IPD). As described in \cite{Axelrod1980a, Knight2016, Press2012} a
number of strategies have been developed to take advantage of the history of
play. Recently, some strategies referred to as zero determinant (ZD) strategies
\cite{Press2012} can manipulate some players through extortionate mechanisms.

@@

\begin{enumerate}
 \item Zero determinant strategies are not particularly effective for $N > 2$
 \item Complex strategies can be effective, and in fact can naturally evolve
     through evolutionary processes to outperform designed strategies.
 \item The strongest resistors specifically evolve or have a handshake mechanism.
 \item Strong invaders are generally cooperative strategies that do not defect
 first but retaliate to varying degrees of intensity against strategies that defect.
 \item Strategies evolved to maximize their total payoff can be strong invaders
 and achieve mutual cooperation with many other strategies.
\end{enumerate}

@@

The Moran process was initially introduced in \cite{Moran1957}. It has since
been used in a variety of settings including the understanding of the spread of
cooperative and non-cooperative behaviour such as cancer \cite{West2016} and the
emergence of cooperative behaviour in spatial topologies \cite{Nowak2017}.
However these works mainly consider relatively simple strategies. Some work has
looked at evolutionary stability of agent-based strategies within the Prisoner's Dilemma
\cite{Li2014} but this is not done in the more widely used setting of the Moran
process, rather in terms of infinite population stability. In \cite{Baek2016}
Moran processes are studied in a theoretical framework for a small subset of
strategies. The subset included memory one strategies: strategies that recall
the events of the previous round only.

@@

All of the training code is archived at \cite{marc_harper_2017_824264}. This
software is (similarly to the Axelrod library) available on github
\url{https://github.com/Axelrod-Python/axelrod-dojo} with documentation to
train new strategies easily. Training
typically takes less than 100 generations and can be completed within several
hours on commodity hardware.

@@

Section~\ref{sec:validation} will further validate the methodology by comparing
simulated results to analytical results in some cases. The main results of this
manuscript are presented in Section~\ref{sec:empirical_results} which will
present a detailed analysis of all the data generated. Finally,
Section~\ref{sec:conclusion} will conclude and offer future avenues for the work
presented here.

@@

Using this it is a known result \cite{Nowak2017} that the fixation probability
of the first strategy in a population of \(i\) individuals of the first type
and \(N-i\) individuals of the second:

@@

Comparisons of \(x_1, x_{N/2}, x_{N-1}\) are shown in
Figure~\ref{fig:comparison_deterministic}. The points represent the simulated
% TODO Add description of lines shown.
values and the line shows the theoretical value. Note that these are all
deterministic strategies and show a perfect match between the expected value
of (\ref{equ:fixation_probability}) and the actual Moran process for all
strategy pairs. Figure~\ref{fig:comparison_stochastic} shows the fixation probabilities for
stochastic strategies. These are no longer a good match which highlights the
weakness of assuming a given interaction between two IPD strategies can be
summarised with a set of utilities as shown in
(\ref{equ:payoff_matrix}). For any given pair of strategies it is possible to
obtain \(p_{i,i-1}, p_{i,i+1}, p_{ii}\) exactly (as opposed to the
approximations offered by (\ref{equ:p_up}), (\ref{equ:p_down}) and
(\ref{equ:p_stay})). Obtaining these requires particular analysis for a given
pair and can be quite a complex endeavour for stochastic strategies with long
memory: this is not necessary for the purposes of this work.
All data generated for this validation exercise can be found
at \cite{data}.
% TODO Include a statistical test (Bonferroni)

@@

\begin{enumerate}
    \item The top strategy is the Collective Strategy (CS) which has a simple
        handshake mechanism described above.
    \item Defector: it always defects. Since it has no interactions with other
        defectors (recall that \(N=2\)), its aggressiveness is rewarded.
    \item Aggravater, which plays like Grudger (responding to any
        defections with unconditional defections throughout) however starts by
        playing 3 defections.
    \item Predator, a finite state machine described in \cite{Ashlock2006}.
    \item Handshake, a slightly less aggressive version of the Collective
        Strategy \cite{robson1989}. As long as the initial sequence is played
        then it cooperates. Thus it will do well in a population consisting of
        many members of itself just as the Collective Strategy does. The
        difference is that CS will defect after the handshake if the opponent
        defects while handshake will not.
\end{enumerate}

@@

The top 16 strategies are given in Tables~\ref{tbl:top_invade}.

@@

There are are only two new strategies that appear in the top ranks for
\(x_{N-1}\): TF1 and TF2. These two strategies are with CS the strongest
resistors. They all have handshakes, and whilst the handshakes of CS and
Handshake (which ranks highly for the smaller values of \(N\)) were
programmed, the handshakes of TF1 and TF2 evolved through an evolutionary
process without any priming.

@@

It is evident through
Sections~\ref{sec:two_individuals},~\ref{sec:strong_invaders}
and~\ref{sec:strong_resistors} that performance of strategies not only depends
on the initial population distribution but also that there seems to be a
difference depending on whether or not \(N>2\). This will be explored further in
the next section, looking not only at \(x_1\) and \(x_{N-1}\) but also consider
\(x_{N/2}\).

@@

\begin{table}[!hbtp]
    \centering
    \scriptsize
    \scalebox{0.7}{\input{tex/change_of_rank_invade.tex}}
    \caption{Invasion: Fixation ranks of some strategies according to \(x_1\) for different
    population sizes}
    \label{tbl:ranks_v_size_invade}
\end{table}

\begin{table}[!hbtp]
    \centering
    \scriptsize
    \scalebox{0.7}{\input{tex/change_of_rank_resist.tex}}
    \caption{Resistance: Fixation ranks of some strategies according to \(x_{N-1}\) for different
    population sizes}
    \label{tbl:ranks_v_size_resist}
\end{table}

@@

\begin{table}[!hbtp]
    \centering
    \scriptsize
    \input{tex/change_of_rank_coexist.tex}
    \caption{Ranks of some strategies according to \(x_{N/2}\) for different
    population sizes}
    \label{tbl:ranks_v_size_coexist}
\end{table}

Figure~\ref{fig:correlation_coefficients} show the correlation coefficients
of the ranks of strategies in differing population size. How well a strategy
performs in any Moran process for \(N>2\) has
little to do with the performance for \(N=2\). This illustrates why the strong
performance of zero determinant strategies predicted in \cite{Press2012} does
not extend to larger populations. This was discussed theoretically in
\cite{Adami2013} and observed empirically in these simulations.

@@

While TF1 and TF2 are competent invaders, the best invaders
in the study do not appear to employ strict handshakes, and are generally
cooperative strategies. TF3, which does not use a handshake, is a better invader
than TF1 and TF2 but not as good a resistor. Nevertheless it was the result
of the same kind of training procceses and is a better combined invader-resistor
than the invaders that were trained previously to maximize payout.

@@

A detailed empirical analysis of 164 strategies of the IPD within a pairwise
Moran process has been carried out. All \(\binom{164}{2}=13,366\) possible
ordered pairs of strategies have been placed in a Moran process with different
starting values allowing the each strategy to attempt to invade the other.
This is the largest such experiment carried out and has led to many insights.

>>>>

% Title
\title{Evolution Reinforces Cooperation with the Emergence of Self-Recognition
       Mechanisms: an Empirical Study of the Moran process for the Iterated
       Prisoner's Dilemma}
       % TODO Possibly change
\author[1]{Vincent Knight}
\author[2]{Marc Harper}
\author[1]{Nikoleta E. Glynatsi}
\author[3]{Owen Campbell}

@@

The Prisoner's Dilemma (PD) \cite{Flood1958} is a fundamental two player game
used to model a variety of strategic interactions. Each player chooses
simultaneously and independently
between cooperation (C) or defection (D). The payoffs of
the game are defined by the matrix $\begin{pmatrix} R & S \\ T & P
\end{pmatrix}$, where $T > R > P > S$ and $2R > T + S$. The PD is a one
round game, but is commonly studied in a manner where the prior outcomes
matter. This repeated form is called the Iterated Prisoner's
Dilemma (IPD). As described in \cite{Axelrod1980a, Knight2016, Press2012} a
number of strategies have been developed to take advantage of the history of
play. Recently, strategies referred to as zero determinant (ZD) strategies
\cite{Press2012} can manipulate some players through extortionate mechanisms.

@@

\begin{enumerate}
 \item Zero determinant strategies are not particularly effective for $N > 2$
 \item Complex strategies can be effective, and in fact can naturally evolve
     through evolutionary processes to outperform designed strategies.
 \item The strongest resistors specifically evolve or possess a handshake mechanism.
 \item Strong invaders are generally cooperative strategies that do not defect
 first but retaliate to varying degrees of intensity against strategies that defect.
 \item Strategies evolved to maximize their total payoff can be strong invaders
 and achieve mutual cooperation with many other strategies.
\end{enumerate}

@@

The Moran process was initially introduced in \cite{Moran1957}. It has since
been used in a variety of settings including the understanding of the spread of
cooperative and non-cooperative behaviour such as cancer \cite{West2016} and the
emergence of cooperative behaviour in spatial topologies \cite{Nowak2017}.
However these works mainly consider relatively simple strategies. A few works 
looked at evolutionary stability of agent-based strategies within the Prisoner's Dilemma
\cite{Li2014} but this is not done in the more widely used setting of the Moran
process, rather in terms of infinite population stability. In \cite{Baek2016}
Moran processes are studied in a theoretical framework for a small subset of
strategies. The subset included memory one strategies: strategies that recall
the events of the previous round only.

@@

All of the training code is archived at \cite{marc_harper_2017_824264}. This
software is (similarly to the Axelrod library) available on 
github\footnote{\url{https://github.com/Axelrod-Python/axelrod-dojo}} 
with documentation to
train new strategies easily. Training
typically takes less than 100 generations and can be completed within several
hours on commodity hardware.

@@

Section~\ref{sec:validation} will further validate the methodology by comparing
simulated results to analytical results in a few selected cases. The main 
results of this
manuscript are presented in Section~\ref{sec:empirical_results} which will
present a detailed analysis of all the data generated. Finally,
Section~\ref{sec:conclusion} will conclude and offer future avenues for the work
presented here.

@@

Using this the fixation probability
of the first strategy in a population of \(i\) individuals of the first type
and \(N-i\) individuals of the second, is given by \cite{Nowak2017}:

@@

Comparisons of \(x_1, x_{N/2}, x_{N-1}\) are shown in
Figure~\ref{fig:comparison_deterministic}. The points represent the simulated
% TODO Add description of lines shown.
values and the line shows the theoretical value. Note that these are all
deterministic strategies and show a perfect match between the expected value
of (\ref{equ:fixation_probability}) and the actual Moran process for all
strategy pairs. Figure~\ref{fig:comparison_stochastic} shows the fixation probabilities for
stochastic strategies. These are no longer a good match which highlights the
% TODO Rephrase this sentence: possibly when doing the statistical test
weakness of assuming a given interaction between two IPD strategies can be
summarised with a set of utilities as shown in
(\ref{equ:payoff_matrix}). For any given pair of strategies it is possible to
obtain \(p_{i,i-1}, p_{i,i+1}, p_{ii}\) exactly (as opposed to the
approximations offered by (\ref{equ:p_up}), (\ref{equ:p_down}) and
(\ref{equ:p_stay})). Obtaining these requires particular analysis for a given
pair and can be quite a complex endeavour for stochastic strategies with long
memory: this is not necessary for the purposes of this work.
All data generated for this validation exercise can be found
at \cite{data}.
% TODO Include a statistical test (Bonferroni)

@@

\begin{enumerate}
    \item The top strategy is the Collective Strategy (CS) which has a simple
        handshake mechanism described above.
    \item Defector: it always defects. Since it has no interactions with other
        defectors (recall that \(N=2\)), its aggressiveness is rewarded.
    \item Aggravater, which plays like Grudger (responding to any
        defections with unconditional defections throughout) however starts by
        playing 3 defections.
    \item Predator, a finite state machine described in \cite{Ashlock2006}.
    \item Handshake, a slightly less aggressive version of the Collective
        Strategy \cite{robson1989}. As long as the initial sequence is played
        then it cooperates. Thus it will do well in a population consisting of
        many members of itself just as the Collective Strategy does. The
        difference is that CS will defect after the handshake if the opponent
        defects while Handshake will not.
\end{enumerate}

@@

The top 16 strategies are given in Table~\ref{tbl:top_invade}.

@@

There are only two new strategies that appear in the top ranks for
\(x_{N-1}\): TF1 and TF2. These two strategies are with CS the strongest
resistors. They all have handshakes, and whilst the handshakes of CS and
Handshake (which ranks highly for the smaller values of \(N\)) were
programmed, the handshakes of TF1 and TF2 evolved through an evolutionary
process without any priming.

@@

It is evident through
Sections~\ref{sec:two_individuals},~\ref{sec:strong_invaders}
and~\ref{sec:strong_resistors} that performance of strategies not only depends
on the initial population distribution but also that there seems to be a
difference depending on whether or not \(N>2\). This will be explored further in
the next section, looking not only at \(x_1\) and \(x_{N-1}\) but also
considering
\(x_{N/2}\).

@@

\begin{table}[!hbtp]
    \centering
    \scriptsize
    \scalebox{0.7}{\input{tex/change_of_rank_invade.tex}}
    \caption{Invasion: Fixation ranks of a few selected strategies according to \(x_1\) for different
    population sizes}
    \label{tbl:ranks_v_size_invade}
\end{table}

\begin{table}[!hbtp]
    \centering
    \scriptsize
    \scalebox{0.7}{\input{tex/change_of_rank_resist.tex}}
    \caption{Resistance: Fixation ranks of a few selected strategies according to \(x_{N-1}\) for different
    population sizes}
    \label{tbl:ranks_v_size_resist}
\end{table}

@@

\begin{table}[!hbtp]
    \centering
    \scriptsize
    \input{tex/change_of_rank_coexist.tex}
    \caption{Ranks of a few selected strategies according to \(x_{N/2}\) for different
    population sizes}
    \label{tbl:ranks_v_size_coexist}
\end{table}

Figure~\ref{fig:correlation_coefficients} shows the correlation coefficients
of the ranks of strategies in differing population size. How well a strategy
performs in any Moran process for \(N>2\) has
little to do with the performance for \(N=2\). This illustrates why the strong
performance of zero determinant strategies predicted in \cite{Press2012} does
not extend to larger populations. This was discussed theoretically in
\cite{Adami2013} and observed empirically in these simulations.

@@

While TF1 and TF2 are competent invaders, the best invaders
in the study do not appear to employ strict handshakes, and are generally
cooperative strategies. TF3, which does not use a handshake, is a better invader
than TF1 and TF2 but not as good a resistor. Nevertheless it was the result
of the same kind of training processes and is a better combined invader-resistor
than the invaders that were trained previously to maximize payout.

@@

A detailed empirical analysis of 164 strategies of the IPD within a pairwise
Moran process has been carried out. All \(\binom{164}{2}=13,366\) possible
ordered pairs of strategies have been placed in a Moran process with different
starting values allowing each strategy to attempt to invade the other.
This is the largest such experiment carried out and has led to many insights.
